---
title: "P8105 Homework 2"
author: "Laylah Jones"
date: 2023-09-28
output: github_document
---

# Problem 1

Importing libraries

```{r}
library(tidyverse)
library(readxl)
library(dplyr)
library(tidyr)
```


### Part 1. First we will read and clean the pols-month.csv data.

```{r}

pols_df = 
  read.csv("./fivethirtyeight_datasets/pols-month.csv") |>
  janitor::clean_names() |>
  separate(mon, c("year", "month", "day"), convert = TRUE) |>
  mutate(
    month = case_when(
      month == 1 ~ "January",
      month == 2 ~ "February",
      month == 3 ~ "March",
      month == 4 ~ "April",
      month == 5 ~ "May",
      month == 6 ~ "June",
      month == 7 ~ "July",
      month == 8 ~ "August",
      month == 9 ~ "September",
      month == 10 ~ "October",
      month == 11 ~ "November",
      month == 12 ~ "December"
    )
  ) |>
  mutate(
    president = case_when(
      prez_dem == 1 ~ "dem",
      prez_gop == 1 ~ "gop"
    )
  ) |>
  select(-prez_dem, -prez_gop, -day)
```

### Part 2. Now we will read and clean snp.csv data.
```{r}
snp_df = 
  read.csv("./fivethirtyeight_datasets/snp.csv") |>
  janitor::clean_names() |>
  separate(date, c("month", "day", "year"), convert = TRUE) |>
  mutate(
    year = as.numeric(year),
    year=ifelse(year<=20,year+2000,year+1900),
    month = case_when(
      month == 1 ~ "January",
      month == 2 ~ "February",
      month == 3 ~ "March",
      month == 4 ~ "April",
      month == 5 ~ "May",
      month == 6 ~ "June",
      month == 7 ~ "July",
      month == 8 ~ "August",
      month == 9 ~ "September",
      month == 10 ~ "October",
      month == 11 ~ "November",
      month == 12 ~ "December"
    )
  )|>
  select(-day) |>
  select(year, month, close)
```

### Part 3. Lastly we will read and clean unemployment.csv data.
```{r}
unemployment_df = 
  read.csv("./fivethirtyeight_datasets/unemployment.csv") |>
  janitor::clean_names() |>
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemployment rate"
  ) |>
  mutate(
    month = case_when(
      month == "jan" ~ "January",
      month == "feb" ~ "February",
      month == "mar" ~ "March",
      month == "apr" ~ "April",
      month == "may" ~ "May",
      month == "jun" ~ "June",
      month == "jul" ~ "July",
      month == "aug" ~ "August",
      month == "sep" ~ "September",
      month == "oct" ~ "October",
      month == "nov" ~ "November",
      month == "dec" ~ "December"
    )
  ) 
```

### Part 4. Merge data sets

```{r}
pols_snp_merge = 
  left_join(pols_df, snp_df, by = c("year", "month"))

merge_df = 
  left_join(pols_snp_merge, unemployment_df, by = c("year", "month"))
```

### Part 5. Dataset explaination and description

The pols_df originally contained information about the presidency and the number of national politicians who were either democratic or republican at the time from 1947 to 2015, and was structured with a variable labelled "mon" signifying the date in the format of "YYYY-MM-DD". By cleaning the data, the "mon" variable was separated into integer variables for "year", "month", and "day". The "month" variable was then mutated from a numeric variable to the corresponding month name. Additionally, a "president" variable was created using the variables "gop" for Republican presidents and "dem" for Democratic presidents, while removing the "prez_dem" and "prex_gop" variables. The resulting pols_df dataset includes dimensions of 822 observed rows by 9 columns and key variables including "year", "month", "president" and other variables related to presidency.

The snp_df originally contained information about the Standard & Poor stock market index, with columns for "year", "month", "day", and "close" values at the time from 1950 to 2015. Cleaning consisted of arranging the data by "year" and "month" to align with the other datasets, and changing the year from "YY" to "YYYY". The dataset includes dimensions of 787 observed rows and 3 columns.

The unemployment_df originally contained information about the percentage of unemployment from 1948 to 2015 in a "wide" format. Cleaning involved converting the dataset from a "wide" to "long" format. Key variables were mutated by changing it from lowercase to uppercase to align with the pols_df. The dataset includes dimensions of 610 observed rows by 3 columns. (6 rows were NA). 

The merge_df dataset was created by merging the three datasets above, using the left_join function. It includes dimensions of 822 observed rows and 11 columns. Variables include month, year, information of the presidency, number of national politicians who are democratic or republican, closing value of s&p stock index, and the percentage of unemployment at a given time, the entire dataset ranges from 1947 to 2015. Key variables include "President" that signifies the presidents' political party, "Close", and "Unemployment rate".


# Problem 2

### Part 1. read and clean Mr. Trash Wheel data

```{r}
Mr_TrashWheel_df = 
  read_excel("./Mr Trash Wheel Data.xlsx", range = cell_cols("A:N"), sheet = "Mr. Trash Wheel") |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(name = "Mr Trash Wheel") |>
  mutate(homes_powered, homes_powered = weight_tons*500/30)
```

### Part 2. read and clean Professor. Trash Wheel data

```{r}
Prof_TrashWheel_df = 
  read_excel("./Mr Trash Wheel Data.xlsx", range = cell_cols("A:N"), sheet = "Professor Trash Wheel") |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(name = "Professor Trash Wheel") |>
  mutate(homes_powered, homes_powered = weight_tons*500/30)
```

### Part 3. read and clean Gwynnda data




